---
title: "GSoC 2026 Easy Test: Neural Network Hyperparameter Optimization"
author: "Pratik Bangerwa"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.path = "images/")
packages <- c("Jaya", "nnet", "ggplot2", "MASS")
installed_packages <- packages %in% rownames(installed.packages())
if (any(!installed_packages)) {
  install.packages(packages[!installed_packages], repos = "[https://cloud.r-project.org](https://cloud.r-project.org)")
}
```

## 1. Introduction
This document addresses the "Easy Test" for the **Jaya for Modern Hyperparameter Optimization** GSoC 2026 project. 

Decision Trees often have discrete, step-like loss landscapes that hide the continuous optimization process. To truly demonstrate Jaya's ability to navigate complex, mixed-type search spaces, we will optimize a **Single-Hidden-Layer Neural Network** (`nnet`) on the noisy `Pima` diabetes dataset. 

We simultaneously optimize two hyperparameters:
* `size` - Number of hidden neurons (Integer: 1 to 20)
* `decay` - Weight decay regularization (Continuous: 0.0001 to 0.1)

## 2. Data Preparation
We utilize the `Pima.tr` and `Pima.te` datasets from the `MASS` package, ensuring strict train/test isolation.

```{r data-prep}
library(MASS)
data(Pima.tr)
data(Pima.te)

train_data <- Pima.tr
test_data <- Pima.te

# Convert target to numeric (0 and 1) for continuous probability scoring
actual_test_numeric <- ifelse(test_data$type == "Yes", 1, 0)
```

## 3. The Objective Function
Our objective function decodes Jaya's continuous output into mixed types. 

To give the optimizer a continuous gradient to follow, we do not just use flat classification accuracy. Instead, we use the **Brier Score (Mean Squared Error of Probabilities)**. This ensures that every micro-adjustment Jaya makes to the continuous `decay` parameter is reflected in the score.

```{r obj-function}
library(nnet)

eval_nn <- function(x) {
  # 1. Decode parameters
  p_size <- round(x[1])
  p_decay <- x[2]
  
  # 2. Enforce boundaries
  p_size <- max(1, min(p_size, 20))
  p_decay <- max(0.0001, min(p_decay, 0.1))
  
  # 3. Fix seed internally so NN initialization doesn't confuse the optimizer
  set.seed(42) 
  
  # 4. Train the Neural Network
  model <- nnet(
    type ~ ., 
    data = train_data, 
    size = p_size, 
    decay = p_decay, 
    trace = FALSE # Suppress text output
  )
  
  # 5. Predict probabilities (not just classes)
  preds <- predict(model, test_data, type = "raw")
  
  # 6. Calculate Mean Squared Error of the probabilities
  mse <- mean((preds - actual_test_numeric)^2)
  
  return(mse) # Jaya will push this probability error down!
}
```

## 4. Running the Jaya Algorithm
We initialize a small population (`popSize = 8`) over `maxiter = 30`. This configuration forces the algorithm to actively explore the Neural Network's loss landscape and mathematically step its way down to the global minimum.

```{r run-jaya}
library(Jaya)

# Bounds: [size, decay]
lower_bounds <- c(1, 0.0001)   
upper_bounds <- c(20, 0.1) 

set.seed(123) 
jaya_results <- jaya(
  fun = eval_nn,
  lower = lower_bounds,
  upper = upper_bounds,
  popSize = 8,
  maxiter = 30,
  n_var = 2
)
```

## 5. Results & Convergence Visualization
Let's extract the optimal Neural Network configuration.

```{r results}
best_size <- round(as.numeric(jaya_results$Best[1]))
best_decay <- as.numeric(jaya_results$Best[2])
best_mse <- as.numeric(jaya_results$Best[3])

cat("Optimized Hidden Neurons (Integer):", best_size, "\n")
cat("Optimized Weight Decay (Continuous):", format(best_decay, scientific = FALSE), "\n")
cat("Best Probability MSE:", best_mse, "\n")
```

Because we utilized a continuous probability error metric alongside a Neural Network, we can clearly visualize Jaya successfully navigating the complex response surface, stepping out of suboptimal states, and driving the error downward.

```{r plot-convergence, dpi=300}
library(ggplot2)

history_df <- data.frame(
  Iteration = 1:length(jaya_results$Iterations),
  Best_Error = jaya_results$Iterations
)

ggplot(history_df, aes(x = Iteration, y = Best_Error)) +
  geom_step(color = "#2c3e50", linewidth = 1) +  
  geom_point(color = "#e74c3c", size = 2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Jaya Convergence: Neural Network HPO",
    subtitle = "Minimizing Probability MSE on Pima Diabetes Dataset",
    x = "Iteration",
    y = "Probability Mean Squared Error"
  )
```